---
title: "Loops and Iteration"
author: "Jessica French"
date: "2022-08-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

```
```{r}
library(tidyverse)
library(palmerpenguins)
```

# `if` statements 

```{r}
burrito <- 2.4

if (burrito > 3) {
  print("I love burritos!")
} 
```

```{r}
my_ships <- c("Millenium Falcon", "X-wing", "Tie_fighter", "Death Star")

str_detect(my_ships, pattern = "r") #goes through every value of the vector and checks for the value specified in pattern = 

phrase <- "I love burritos"
#can use any function in the if statement that returns TRUE or FALSE
if (str_detect(phrase, "love")) {
  print("I am a burrito fan.")
}
```
## `if-else` statement
```{r}
pika <- 89.1

if (pika > 60) {
  print("mega pike")
  } else {
  print("Normal pika")
  }

```

```{r}
marmot <- 2.8

if (marmot < 0.5) {
  print("this is a small marmot")
} else if (marmot >= 0.5 & marmot < 3) {
  print("this is a medium marmot")
} else if (marmot > 3) {
  print("this is a large marmot")
}

```



```{r}
tatooine_characters <- starwars |> 
  filter(homeworld == "Tatooine")

tatooine_characters

#Use %in% operator to check for multiple positive matches

tatooine_endor_naboo <- starwars |> 
  filter(homeworld %in% c("Tatooine", "Naboo", "Endor"))

#In words: does the vale in the homeworld variable for that row exist IN that vector c("Tatooine, "Naboo", "Endor)
```

```{r}
git_pumped <- {function()
  print("You got this!!")}
```
# Syntactic sugar
- i.e. lets not reinvent the wheel
  - apply() in base R
  - dplyr:: across()
  - {purrr} package, purrr.tidyverse.org for more info
  
# Example group_by() %>% summarize() to get summary stats for a group that you specify
```{r}
#mean bill length by speceis

penguins |> #in penguins dataset
  group_by(species, island) |>  #creates groups based on the species column also assumes that everything you do after this will be by group
  summarize(mean_bill_length = mean(bill_length_mm, na.rm = TRUE),
            sd_bill_length = sd(bill_length_mm, na.rm = TRUE)) #na.rm = TRUE removes na values so that it doesnt return na for the mean 

#Find the mean and standard deviation of height in starwars grouped by homeworld

starwars |> 
  group_by(homeworld) |> 
  summarize(mean_height = mean(height, na.rm = TRUE),
            sd_height = sd(height, na.rm = TRUE))


#Fidn teh max, min for flipper length, and sample siize fo each species of penguins

penguins |> 
  group_by(species, sex) |> 
  summarize(min_flipper = min(flipper_length_mm, na.rm = TRUE),
            max_flipper = max(flipper_length_mm, na.rm = TRUE),
            sample_size = n())
  
#application of across
penguins |> 
  group_by(species, sex) |> 
  summarize(across(where(is.numeric), mean, na.rm = TRUE)) #can also give a vector of function to do. Can also pass a vector of column names rather than a conditional that applys to the whole data frame. 
```

## A for loop with cumulative calculations

```{r}
tree_height <- c(1, 2, 6, 10, 14, 20, 45, 30, 12)

for (i in seq_along(tree_height)) {
  val = tree_height[i] + tree_height[i + 1]
  print(val)
}

```
## A for loop with a conditional statement 

```{r}
animal_ages <- vector(mode = "numeric", length = length(species))
species <- c("dog", "elephant", "goat", "dog", "dog", "elephant")
age_human <- c(3, 8, 4, 6, 12, 18)

for (i in seq_along(species)) {
  if (species[i] == "dog") {
    animal_age <- age_human[i] * 7
  } else if (species[i] == "elephant") {
    animal_age <- age_human[i] * 0.88
  } else if (species[i] == "goat") {
    animal_age <- age_human[i] * 4.7
  }
  animal_ages[i] <- animal_age
}

```

#For loops across data frames
```{r}
mean_mtcars <- vector(mode = "numeric", length = ncol(mtcars))

for (i in seq_along(mtcars)) {
  mean_val <- mean(mtcars[[i]], na.rm = TRUE)
  mean_mtcars[i] <- mean_val
}

mean_mtcars
```

## Functional programming examples

adding syntactic sugar (apply, across, map_df)

```{r}
apply(X = mtcars, MARGIN = 2, FUN = mean)
```

```{r}
map(.x = mtcars, .f = mean)

map_df(.x = mtcars, .f = mean)
```

